Queue Aware Precoder Design for Space Frequency Resource Allocation

### 2014-09-09 Report/sub_channel_wsrm.tex
revision - 1

The complexity involved in the \ac{JSFRA} scheme scales significantly with the increase in the number of sub-channels considered in the formulation. In addition to the increased complexity, the rate of convergence to the optimal precoders also degrades due to its dependency on the problem size. In order to mitigate this, we provide an alternative sub-optimal solution, in which the precoders are designed over each sub-channel independently in a sequential manner by taking the remaining number of queued bits in the formulation. It provides a sub-optimal solution to the optimal way of distributing the sub-channel wise precoder design via primal/dual decomposition methods discussed in \cite{palomar2006tutorial,boyd2011distributed}.

The proposed \acf{QM} \ac{SRA} scheme decouples the problem by fixing the power across each sub-channel to a constant value \me{P_{\max,n}} as compared to the global power constraint defined by \eqref{eqn-4.3}. In contrast to the decomposition based approach for the sub-channel wise resource allocation, where the primal/dual variables are exchanged, this method requires the update on the number of queued bits before each sub-channel wise optimization. The total number of queued bits for each user are updated by the difference between the total number of queued bits present during the current slot to the total number of bits that are guaranteed by the earlier sub-channel allocations for the same slot as
\begin{equation}
Q_{k,n} = \max{\Big \lbrace Q_k - \sum_{j = 1}^{n-1} \, \sum_{l = 1}^{L} \, t_{l,k,j} ,0 \Big \rbrace }, \; \forall \; k \in \mathcal{U}
\label{eqn-weight}
\end{equation}
where \me{Q_{k,n}} is the total number of queued bits used in the optimization problem carried out for the sub-channel \me{n}. In the expression \eqref{eqn-weight}, \me{Q_k} denotes the total number of queued bits waiting to be transmitted for the user \me{k} during the current slot and \me{t_{l,k,j}} is the rate or guaranteed bits allocated over the sub-channel \me{j}. The current scheme is very sensitive to the order in which the sub-channels are selected for the optimization problem. The algorithmic representation of the \acf{QM} \ac{SRA} scheme is shown in Algorithm \ref{algo-5}
\begin{algorithm}
 \SetAlgoLined
 \DontPrintSemicolon
 \BlankLine
 \SetKwInput{KwInit}{Initialize}
 \KwIn{\me{a_k, \, Q_k, \, \mvec{H}{b,k,n},\; \fall b \in \mathcal{B}, \, \fall k \in \mathcal{U}}}
 \KwIn{permute \me{\mathcal{N} \rightarrow \tilde{\mathcal{N}}}}
 \For{\me{n \leftarrow 1 } \KwTo \me{N}}{
 update \me{Q_{k,n}} using \eqref{eqn-weight} and let \me{\hat{n} = \tilde{\mathcal{N}}(n)}. \;
 \KwOut{\me{\mvec{m}{l,k,\hat{n}}} and \me{\mvec{w}{l,k,\hat{n}} \fall l \in \set{1,2,\dotsc,L}}}
 \KwInit{\me{i=0}, \me{j=0} and the transmit precoders \me{\tilde{\mbf{m}}_{l,k,\hat{n}}} randomly satisfying per sub-channel power constraint \me{P_{\max,\hat{n}}}}
 update \me{\mvec{w}{l,k,\hat{n}}} with \eqref{eqn-10} and \me{\tilde{\mbf{u}}_{l,k,\hat{n}}} with \eqref{eqn-8} using \me{\tilde{\mbf{m}}_{l,k,\hat{n}}}.\;
 \Repeat{Queue convergence or \me{j \geq J_{\max}}}{
 \Repeat{\ac{SCA} convergence or \me{i \geq I_{\max}}}{
 solve for the transmit precoders \me{\mvec{m}{l,k,\hat{n}}} using \eqref{eqn-9} \;
 update the constraint set \eqref{eqn-8} with \me{\tilde{p}_{l,k,\hat{n}}}, \me{\tilde{q}_{l,k,\hat{n}}} and \me{\tilde{\beta}_{l,k,\hat{n}}} using \eqref{eqn-wsrm-expr} with the precoders \me{\mvec{m}{l,k,\hat{n}}} obtained from the previous step. \;
 $i = i + 1$. \;
 }
 update the receive beamformers \me{\mvec{w}{l,k,n}} using \eqref{eqn-10} with the recent precoders \me{\mvec{m}{l,k,n}}. \;
 $j = j + 1$. \;
 }
 }
 \caption{Algorithm of \acs{SRA} scheme}
 \label{algo-5}
\end{algorithm}

>>>>

The complexity involved in the \ac{JSFRA} scheme scales significantly with the increase in the number of sub-channels considered in the formulation. In addition to the increased complexity, the rate of convergence to the optimal precoders also degrades due to its dependency on the problem size. In order to mitigate this, we provide an alternative sub-optimal solution, in which the precoders are designed over each sub-channel independently in a sequential manner by taking the remaining number of queued bits in the formulation. The centralized precoder design can also be performed across each sub-channel by using the decomposition methods discussed in \cite{palomar2006tutorial,boyd2011distributed}.

The proposed \acl{QM} \ac{SRA} formulation enables us to solve for the transmit precoders of all the users associated with the coordinating \ac{BS} in the set \me{\mathcal{B}} over each sub-channel independently by fixing the transmit power on each sub-channel to a constant value \me{P_{\max,n}} as compared to the global power constraint defined by \eqref{eqn-4.3}. In contrast to the decomposition based approach for the sub-channel wise resource allocation, where the primal/dual variables are exchanged, this method requires the update on the number of queued bits before each sub-channel wise optimization. The total number of queued bits for each user are updated by the difference between the total number of queued bits present during the current slot to the total number of bits that are guaranteed by the earlier sub-channel allocations for the same slot as
\begin{equation}
Q_{k,n} = \max{\Big \lbrace Q_k - \sum_{j = 1}^{n-1} \, \sum_{l = 1}^{L} \, t_{l,k,j} ,0 \Big \rbrace }, \; \forall \; k \in \mathcal{U}
\label{eqn-weight}
\end{equation}
where \me{Q_{k,n}} is the total number of queued bits used in the optimization problem carried out for the sub-channel \me{n}. In the expression \eqref{eqn-weight}, \me{Q_k} denotes the total number of queued bits waiting to be transmitted for the user \me{k} during the current slot and \me{t_{l,k,j}} is the rate or guaranteed bits allocated over the sub-channel \me{j}. The current scheme is sensitive to the order in which the sub-channels are selected for the optimization problem. The algorithmic representation of the \ac{SRA} scheme is shown in Algorithm \ref{algo-5}
\begin{algorithm}
 \SetAlgoLined
 \DontPrintSemicolon
 \BlankLine
 \SetKwInput{KwInit}{Initialize}
 \KwIn{\me{a_k, \, Q_k, \, \mvec{H}{b,k,n},\; \fall b \in \mathcal{B}, \, \fall k \in \mathcal{U}}}
 \KwIn{permute \me{\mathcal{N} \rightarrow \tilde{\mathcal{N}}}}
 \For{\me{n \leftarrow 1 } \KwTo \me{N}}{
 update \me{Q_{k,n}} using \eqref{eqn-weight} and let \me{\hat{n} = \tilde{\mathcal{N}}(n)}. \;
 \KwOut{\me{\mvec{m}{l,k,\hat{n}}} and \me{\mvec{w}{l,k,\hat{n}} \fall l \in \set{1,2,\dotsc,L}}}
 \KwInit{\me{i=0}, \me{j=0} and the transmit precoders \me{\tilde{\mbf{m}}_{l,k,\hat{n}}} randomly satisfying per sub-channel power constraint \me{P_{\max,\hat{n}}}}
 update \me{\mvec{w}{l,k,\hat{n}}} with \eqref{eqn-10} and \me{\tilde{\mbf{u}}_{l,k,\hat{n}}} with \eqref{eqn-8} using \me{\tilde{\mbf{m}}_{l,k,\hat{n}}}.\;
 \Repeat{Queue convergence or \me{j \geq J_{\max}}}{
 \Repeat{\ac{SCA} convergence or \me{i \geq I_{\max}}}{
 solve for the transmit precoders \me{\mvec{m}{l,k,\hat{n}}} using \eqref{eqn-9} \;
 update the constraint set \eqref{eqn-8} with \me{\tilde{p}_{l,k,\hat{n}}}, \me{\tilde{q}_{l,k,\hat{n}}} and \me{\tilde{\beta}_{l,k,\hat{n}}} using \eqref{eqn-wsrm-expr} with the precoders \me{\mvec{m}{l,k,\hat{n}}} obtained from the previous step. \;
 $i = i + 1$. \;
 }
 update the receive beamformers \me{\mvec{w}{l,k,n}} using \eqref{eqn-10} with the recent precoders \me{\mvec{m}{l,k,n}}. \;
 $j = j + 1$. \;
 }
 }
 \caption{Algorithm of \acs{SRA} scheme}
 \label{algo-5}
\end{algorithm}

### 2014-09-09 Report/system_model.tex
Signed-off-by: Ganesh <gvenkatr@ee.oulu.fi>
modified for review
revision - 1

We consider a downlink \ac{MIMO} \ac{BC} scenario in an \ac{OFDM} framework with \me{N} sub-channels and \me{N_B} \acp{BS} each equipped with \me{N_T} transmit antennas, serving \me{K} users each with \me{N_R} receive antennas. The set of users associated with \ac{BS} \me{b} is denoted by \me{\mathcal{U}_b} and the set \me{\mathcal{U}} represents all users in the system, i.e., \me{\mathcal{U} =\underset{b\in \mathcal{B}}{\cup }\mathcal{U}_b}, where \me{\mathcal{B}} is the set of all coordinating \acp{BS}. Data for user \me{k} is transmitted from only one \ac{BS} which is denoted by \me{b_k\in \mathcal{B}}. We denote by \me{\mathcal{C} = \set{1,2,\dotsc,N}} the set of all sub-channel indices available in the system.

In this paper we adopt linear beamforming technique at \acp{BS}. Specifically, the data symbols \me{d_{l,k,n}} for  user \me{k} on the \me{l^\mathrm{th}} spatial stream over the sub-channel \me{n} is multiplied with the beamformer \me{\mvec{m}{l,k,n} \in \mathbb{C}^{N_T \times 1}} before being transmitted.   In order to detect  multiple spatial streams at the receiver, a receive beamforming vector \me{\mvec{w}{l,k,n}} is employed at each user. Consequently, the received data symbol corresponding to the \me{l^\mathrm{th}} spatial stream over sub-channel \me{n} at user $k$ is given by
\begin{equation}\label{eqn-1}
\hat{d}_{l,k,n} = \mvec{w}{l,k,n}^\herm \mvec{H}{b_k,k,n} \,\mvec{m}{l,k,n} d_{l,k,n} + \mvec{w}{l,k,n}^\herm \sum_{i \in \mc{U} \backslash \set{k}} \mvec{H}{b_i,k,n} \sum_{j = 1}^L \mvec{m}{j,i,n}d_{j,i,n} + \mvec{w}{l,k,n}^\herm \mvec{n}{k,n}
\end{equation}
where \me{\mvec{H}{b,k,n} \in \mathbb{C}^{N_R \times N_T}} is the channel between \ac{BS} \me{b} and  user \me{k} on sub-channel \me{n}, and \me{\mvec{n}{k,n}} \me{\sim \mc{CN}(0,N_0)} is the additive noise vector for the user \me{k} on the \me{n^\mathrm{th}} sub-channel and \me{l^\mathrm{th}} spatial stream. In \eqref{eqn-1}, \me{L = rank(\mvec{H}{b,k,n}) = min(N_T,N_R)} is the maximum number of spatial streams. Assuming independent detection of data streams, we can write the signal-to-interference-plus-noise ratio (SINR) as
\begin{equation}\label{eq:SINR}
\gamma_{l,k,n} = \dfrac{\left |\mvec{w}{l,k,n}^\herm \, \mvec{H}{b_k,k,n} \, \mvec{m}{l,k,n} \right |^2}{N_0\|\mvec{w}{l,k,n}\|^2 + \sum_{(j,i) \neq (l,k)} |\mvec{w}{l,k,n}^\herm \mvec{H}{b_i,k,n} \mvec{m}{j,i,n} |^2}.
\end{equation}
Let \me{Q_k} be the number of backlogged packets which are destined for the user \me{k} at a given scheduling instant. The queue dynamics of the user \me{k} are modeled using the Poisson arrival process with the average packet arrivals of \me{A_k = \mathbf{E}_i\{\lambda_k\}} packets/bits, where \me{\lambda_k(i) \sim \mathrm{Pois}(A_k)} represents the instantaneous number of packets or bits arriving for the user \me{k} at the \me{{i}^\mathrm{th}} instant. The total number of queued packets at the \me{{(i+1)}^\mathrm{th}} instant for the user \me{k}, represented by \me{Q_k(i+1)}, is given by
\begin{equation}
Q_k(i+1) = \Big [ Q_k(i) - t_k(i) \Big ]^+ + \lambda_k(i)
\label{eqn-2a}
\end{equation}
where \me{[x]^+ \equiv \max{ \lbrace x,0  \rbrace}} and \me{t_k} denotes the transmission in bits for user \me{k}. For a \ac{MIMO} \ac{OFDM} system,
\begin{equation}
t_k(i) = \sum_{n = 1}^N \, \sum_{l = 1}^L \, t_{l,k,n}(i)
\end{equation}
where \me{t_{l,k,n}} denotes the transmitted bits over \me{\ith{l}} spatial stream on the \me{\ith{n}} sub-channel. The maximum rate achieved over the \me{(l,n)} space-frequency resource is given by \me{t_{l,k,n} \leq \log_2(1 + \gamma_{l,k,n})} for the \ac{SINR} of \me{\gamma_{l,k,n}}.\footnote{This can be achieved by Gaussian signaling} Note that the units of \me{t_k} and \me{Q_k} are in bits defined per channel use.

>>>>

We consider a downlink \ac{MIMO} \ac{BC} scenario in an \ac{OFDM} framework with \me{N} sub-channels and \me{N_B} \acp{BS} each equipped with \me{N_T} transmit antennas, serving \me{K} users each with \me{N_R} receive antennas. The set of users associated with \ac{BS} \me{b} is denoted by \me{\mathcal{U}_b} and the set \me{\mathcal{U}} represents all users in the system, i.e., \me{\mathcal{U} =\underset{b\in \mathcal{B}}{\cup }\mathcal{U}_b}, where \me{\mathcal{B}} is the set of all coordinating \acp{BS}. Data for user \me{k} is transmitted from only one \ac{BS} which is denoted by \me{b_k\in \mathcal{B}}. We denote by \me{\mathcal{N} = \set{1,2,\dotsc,N}} the set of all sub-channel indices available in the system.

In this paper we adopt linear beamforming technique at \acp{BS}. Specifically, the data symbols \me{d_{l,k,n}} for  user \me{k} on the \me{l^\mathrm{th}} spatial stream over the sub-channel \me{n} is multiplied with the beamformer \me{\mvec{m}{l,k,n} \in \mathbb{C}^{N_T \times 1}} before being transmitted.   In order to detect  multiple spatial streams at the receiver, a receive beamforming vector \me{\mvec{w}{l,k,n}} is employed at each user. Consequently, the received data symbol corresponding to the \me{l^\mathrm{th}} spatial stream over sub-channel \me{n} at user $k$ is given by
\begin{equation}\label{eqn-1}
\hat{d}_{l,k,n} = \mvec{w}{l,k,n}^\herm \mvec{H}{b_k,k,n} \,\mvec{m}{l,k,n} d_{l,k,n} + \mvec{w}{l,k,n}^\herm \sum_{i \in \mc{U} \backslash \set{k}} \mvec{H}{b_i,k,n} \sum_{j = 1}^L \mvec{m}{j,i,n}d_{j,i,n} + \mvec{w}{l,k,n}^\herm \mvec{n}{k,n}
\end{equation}
where \me{\mvec{H}{b,k,n} \in \mathbb{C}^{N_R \times N_T}} is the channel between \ac{BS} \me{b} and  user \me{k} on sub-channel \me{n}, and \me{\mvec{n}{k,n}} \me{\sim \mc{CN}(0,N_0)} is the additive noise vector for the user \me{k} on the \me{n^\mathrm{th}} sub-channel and \me{l^\mathrm{th}} spatial stream. In \eqref{eqn-1}, \me{L = rank(\mvec{H}{b,k,n}) = min(N_T,N_R)} is the maximum number of spatial streams \footnote{\me{L} streams are initialized but after solving the problem, only \me{L_{k,n} \leq L} non-zero data streams are transmitted}. Assuming independent detection of data streams, we can write the signal-to-interference-plus-noise ratio (SINR) as
\begin{equation}\label{eq:SINR}
\gamma_{l,k,n} = \dfrac{\left |\mvec{w}{l,k,n}^\herm \, \mvec{H}{b_k,k,n} \, \mvec{m}{l,k,n} \right |^2}{N_0\|\mvec{w}{l,k,n}\|^2 + \sum_{(j,i) \neq (l,k)} |\mvec{w}{l,k,n}^\herm \mvec{H}{b_i,k,n} \mvec{m}{j,i,n} |^2}
\end{equation}
Let \me{Q_k} be the number of backlogged packets which are destined for the user \me{k} at a given scheduling instant. The queue dynamics of the user \me{k} are modeled using the Poisson arrival process with the average packet arrivals of \me{A_k = \mathbf{E}_i\{\lambda_k\}} packets/bits, where \me{\lambda_k(i) \sim \mathrm{Pois}(A_k)} represents the instantaneous number of packets or bits arriving for the user \me{k} at the \me{{i}^\mathrm{th}} instant. The total number of queued packets at the \me{{(i+1)}^\mathrm{th}} instant for the user \me{k}, represented by \me{Q_k(i+1)}, is given by
\begin{equation}
Q_k(i+1) = \Big [ Q_k(i) - t_k(i) \Big ]^+ + \lambda_k(i)
\label{eqn-2a}
\end{equation}
where \me{[x]^+ \equiv \max{ \lbrace x,0  \rbrace}} and \me{t_k} denotes the transmission in bits for user \me{k}. For a \ac{MIMO}-\ac{OFDM} system,
\begin{equation}
t_k(i) = \sum_{n = 1}^N \, \sum_{l = 1}^L \, t_{l,k,n}(i)
\end{equation}
where \me{t_{l,k,n}} denotes the transmitted bits over \me{\ith{l}} spatial stream on the \me{\ith{n}} sub-channel. The maximum rate achieved over the \me{(l,n)} space-frequency resource is given by \me{t_{l,k,n} \leq \log_2(1 + \gamma_{l,k,n})} for the \ac{SINR} of \me{\gamma_{l,k,n}}.\footnote{This can be achieved by Gaussian signaling} Note that the units of \me{t_k} and \me{Q_k} are in bits defined per channel use.
